---
title: "How parenthood shapes the gender wage gap"
subtitle: "Relative to men without children (ages 25–44)"
author:
  - "Sakura Hu"
  - "Zhanyi Wang"
date: "`r format(as.Date('2025-10-24'), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(arrow)
library(readr)
library(scales)
library(ggrepel)
library(forcats)
library(dplyr)
library(tidyr)
library(ggplot2)
library(grid)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data <- read_parquet("./data/01-cleaned_data/cleaned_census_data.parquet")
```
# Gender inequality still exist
test reference[@one]and[@two]and[@three]and[@four]and[@five]and[@six]!
In recent years, in order to address problems such as gender discrimination and the undervaluation of women’s work, many countries and regions promote equal pay for work of equal value and the economic empowerment of females. (1. ) As one of these regions, Canada has made continuous progress in addressing gender inequality in employment, but the pay gap remains stubbornly persistent. Women with full-time jobs still earn significantly less than men on average, and this gap is even wider for Indigenous women, women of colour, women with disabilities, and others who face systemic and intersectional barriers.(2. ) In this context, an important question arises: Why does the gender pay gap persist even after policies like equal pay for equal work have been implemented? Education and occupation are often cited as explanations, but the data tells a more complex story, which lies at the crossroads of work, gender, and motherhood.

# Educational disparities are not enough to explain the problem:
A common hypothesis attributes wage differences to education and occupation gaps. In general, higher education means a greater chance of entering occupations with higher average salaries(3.), so if most women choose to work early instead of continuing to obtain higher education, they may lose the opportunity to access high-paying occupations. However, this explanation no longer holds. In fact, women are more likely than men to pursue postsecondary studies, and they represent the majority of bachelor’s degree, college certificate and diploma holders(4.).In addition, across all education levels and occupations, women’s earnings are always lower than men’s. At the same level of education, the wage gap exceeds 15%, and even within the same occupation, the smallest gap between men and women is still about 9.1%. Although higher education and occupational choices may help narrow the gender pay gap, it never fully disappears.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# gap by occupation

# 2-digit NAICS sector labels
naics_labs <- c(
  `11` = "Agriculture, forestry, fishing & hunting",
  `21` = "Mining, quarrying, oil & gas",
  `22` = "Utilities",
  `23` = "Construction",
  `31` = "Manufacturing",
  `41` = "Wholesale trade",
  `44` = "Retail trade",
  `48` = "Transportation & warehousing",
  `51` = "Information & culture",
  `52` = "Finance & insurance",
  `53` = "Real estate, rental & leasing",
  `54` = "Professional, scientific & technical",
  `55` = "Management of companies",
  `56` = "Admin & support, waste mgmt",
  `61` = "Educational services",
  `62` = "Health care & social assistance",
  `71` = "Arts, entertainment & recreation",
  `72` = "Accommodation & food services",
  `81` = "Other services (except public admin)",
  `91` = "Public administration"
)


naics_gap <- data %>%
  mutate(NAICS_lab = recode(as.character(NAICS), !!!naics_labs)) %>%
  filter(!is.na(NAICS_lab)) %>%
  group_by(NAICS_lab, Gender) %>%
  summarise(med_wage = median(Wages, na.rm = TRUE), .groups = "drop_last") %>%
  pivot_wider(names_from = Gender, values_from = med_wage) %>%
  mutate(
    gap_pct = (Men - Women) / Men,
    # (Men − Women) / Men
    gap_cat = case_when(
      gap_pct < 0    ~ "Women higher",
      gap_pct < 0.10 ~ "0–10%",
      gap_pct < 0.20 ~ "10–20%",
      TRUE           ~ "≥20%"
    ),
    # set factor order HIGH → LOW so bars run top (largest) to bottom (smallest) after coord_flip()
    NAICS_lab = fct_reorder(NAICS_lab, gap_pct, .desc = TRUE)
  ) %>%
  tidyr::drop_na(gap_pct)

# palette with strong separation (green for negative, oranges/reds for larger gaps)
pal <- c(
  "Women higher" = "#2ca25f",
  "0–10%"       = "#fee8c8",
  "10–20%"      = "#fdbb84",
  "≥20%"        = "#e34a33"
)

# --- single plot object ---
p_naics <- ggplot(naics_gap, aes(
  x = reorder(NAICS_lab, gap_pct),
  y = gap_pct,
  fill = gap_cat
)) +
  geom_col(width = 0.72) +
  geom_text(
    aes(label = percent(gap_pct, accuracy = 0.1)),
    hjust = -0.12,
    size = 3.1,
    colour = "grey20"
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(name = "Gap size", values = pal) +
  labs(
    title   = "Gender pay gap by industry(gai)",
    x       = NULL,
    y       = "Gap percentage",
    caption = "Figure shows the gender pay gap by industry for full-time, full-year workers.Gaps exist across all sectors, ranging roughly \nfrom 9% to 30%. Finance & insurance is among the highest."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor    = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption.position = "plot",
    plot.caption        = element_text(hjust = 0)
  )

p_naics
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# gap by education
# Compute median wages by Education × Gender, then the gap = (Men − Women) / Men
edu_gap <- data %>%
  filter(!is.na(Edu3)) %>%
  group_by(Edu3, Gender) %>%
  summarise(med_wage = median(Wages, na.rm = TRUE), .groups = "drop_last") %>%
  pivot_wider(names_from = Gender, values_from = med_wage) %>%
  mutate(gap_pct = (Men - Women) / Men) %>%
  drop_na(gap_pct) %>%
  # order categories from lowest → highest gap
  mutate(Edu3 = as.character(Edu3)) %>%
  arrange(gap_pct) %>%
  mutate(Edu3 = factor(Edu3, levels = Edu3))

# palette per education level (tweak if you like)
edu_cols <- c(
  "HS or less"          = "#fdbb84",
  "Non-degree postsec"  = "#9ecae1",
  "BA or higher"        = "#a1d99b"
)

ggplot(edu_gap, aes(x = gap_pct, y = Edu3, fill = Edu3)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             colour = "grey55") +
  geom_text(
    aes(label = percent(gap_pct, accuracy = 0.1)),
    hjust = -0.12,
    size = 3.2,
    colour = "grey20"
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = edu_cols, name = "Education") +
  labs(
    title   = "Gender pay gap by education level",
    x       = "Gap percentage",
    y       = NULL,
    caption = "Figure shows the gender pay gap by education level. The gap is largest among those with non-degree postsecondary \ncredentials (25.3%) and smallest among those with a bachelor’s degree or higher (16.8%)."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "right",
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption.position = "plot",
    plot.caption        = element_text(hjust = 0)
  )

```
If education cannot fully explain the difference, what does? To find out, the answer may lie in a more personal yet universal moment, childbirth.

# Turning point at age 30:
When plotting the wage gap between men and women across age, a distinct pattern begins to appear.


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
#2021 gap by age

# Map census AGEGRP codes to labels
age_labels <- c(
  `7` = "18–19",
  `8` = "20–24",
  `9` = "25–29",
  `10` = "30–34",
  `11` = "35–39",
  `12` = "40–44",
  `13` = "45–49",
  `14` = "50–54",
  `15` = "55–59",
  `16` = "60–64"
)

# Prep 2021 data
df21 <- data %>%
  mutate(
    Gender = recode(as.character(Gender), `1` = "Women", `2` = "Men"),
    Age = recode(as.character(AGEGRP), !!!age_labels)
  ) %>%
  filter(Gender %in% c("Women", "Men"), !is.na(Wages), !is.na(Age))

# Compute gap by age for 2021: Gap = (Men − Women) / Men using medians
gap_age <- df21 %>%
  group_by(Age, Gender) %>%
  summarise(med = median(Wages, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Gender, values_from = med) %>%
  filter(!is.na(Men), !is.na(Women)) %>%
  mutate(Age = factor(Age, levels = age_labels),
         gap_pct = (Men - Women) / Men) %>%
  arrange(Age)

# Red highlight point at 25–29
pts_2529     <- gap_age %>% filter(Age == "25–29")
pt_2529_top  <- pts_2529  # used for the ggrepel note

p_2021 <- ggplot(gap_age, aes(Age, gap_pct, group = 1)) +
  geom_line(linewidth = 1.1, color = "black") +
  geom_point(size = 2, color = "black") +
  # red dot + label at 25–29
  geom_point(
    data = pts_2529,
    aes(Age, gap_pct),
    inherit.aes = FALSE,
    size = 3.8,
    color = "red"
  ) +
  geom_text_repel(
    data = pts_2529,
    aes(Age, gap_pct, label = "Mean maternal \nage in Canada)"),
    color = "red",
    segment.color = "red",
    nudge_y = 0.06,
    box.padding = 0.35,
    size = 3.2,
    show.legend = FALSE
  ) +
  # Y axis: 0–30%
  scale_y_continuous(
    limits = c(0, 0.30),
    breaks = seq(0, 0.30, by = 0.05),
    labels = scales::label_percent(accuracy = 1)
  ) +
  labs(
    title = "Canada’s gender pay gap by age (2021)",
    x = "Age group",
    y = "Gap percentage",
    caption = "Figure shows the gender pay gap by age for full-time, full-year workers. The gap narrows in the \nearly 20s, then rises sharply from 25–30 and continues to increase steadily thereafter."
    
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    # center the title
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

p_2021

```

Before age 30, the pay gap between men and women narrows rapidly as both gain experience and move up early in their careers. However, after 30, this trend reverses, the gap begins to widen steadily with age. Coincidentally, age 30 also corresponds to Canada’s average maternal age for first-time mothers. This turning point might suggest that while early career growth helps women catch up to men in earnings, the onset of motherhood often marks a shift. The data seem to tell a familiar story, after becoming mothers, women usually shift the focus of their lives to their children and family, thus losing some opportunities for promotion or salary increase.

It is also important to recognize that childbirth affects women not only through changes in wages, but also through changes in labour participation itself. After becoming mothers, some women temporarily or permanently leave the workforce to take on domestic and caregiving responsibilities. These women are often classified as “not employed”. Therefore, to fully understand the impact of motherhood on economic inequality, it is necessary to examine both groups, women who remain in paid employment and women who exit the labour market after childbirth. By considering both wage differences among working women and unemployment patterns related to family responsibilities, it can capture a more complete picture of how motherhood affects the outcomes of women's career and economies.

# Implications for career women: 
Among employed individuals, the gender gap in earnings widens sharply after parenthood. As shown in Figure, among workers without children, women earn about 13% less than men, but after having children, the gap expands to roughly 21%. This pattern suggests that while a gender pay gap already exists before parenthood, it becomes substantially larger once children enter the picture, which suggests that having children can exacerbate pay inequality in the workplace.

Furthermore, the impact of having children on earnings moves in opposite directions for men and women. Men with children earn about 7% more than men without children, as many employers tend to view fathers as more stable, responsible, and committed. These qualities can translate into higher pay and faster promotions(5.).

In contrast with men, women without children earn about 13% less than men without children, but after having children, women earn around 15% less. The increasing wage gap after having a child may reflect cultural biases toward mothers, which to be seen as a good mother they are expected to prioritize family over career, but to be viewed as a “Good leader,” they must do the opposite. Under such contradiction, many mothers deviate from traditional career paths by reducing their working hours or leaving the workforce indefinitely, all of which negatively affects their chances of career advancement. The starkly different career situations of women and men after having children illustrate that gender stereotypes and unfairness in the workplace still exist, exacerbating the gender wage gap.

# Unemployment due to childbirth:
While the gender pay gap widens around the mean maternal age, the impact of childbirth is also reflected in women’s unemployment patterns. The first figure compares unemployment rates among women with and without children. The dark red line shows that women with children consistently face higher unemployment rates, particularly between ages 25 and 45. This is because, in addition to childbirth, women in this age group are more likely to take on caregiving responsibilities for children under the age of 15 (6). In contrast, women without children experience a much sharper decline in unemployment after their early twenties and maintain relatively stable employment throughout midlife.


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# --- Parent premium/penalty on a single number line --------------------------

pts <- data %>%
  mutate(
    Gender = recode(as.character(Gender), `1`="Women", `2`="Men"),
    ChildStatus = case_when(PKIDS == 1 ~ "Has children",
                            PKIDS == 0 ~ "No children",
                            TRUE ~ NA_character_)
  ) %>%
  filter(AGEGRP %in% 9:12,                      # 25–44
         Gender %in% c("Men","Women"),
         !is.na(ChildStatus),
         !is.na(Wages)) %>%
  group_by(Gender, ChildStatus) %>%
  summarise(med = median(Wages, na.rm = TRUE), .groups = "drop") %>%
  # baseline = Men — No children
  mutate(baseline = med[Gender == "Men" & ChildStatus == "No children"][1],
         gap = (med - baseline) / baseline) %>%
  mutate(
    # aesthetics for the four dots
    dot_col = case_when(
      Gender == "Men"    & ChildStatus == "No children" ~ "#4F83FF", # blue (baseline)
      Gender == "Men"    & ChildStatus == "Has children" ~ "#2E5BD4", # darker blue
      Gender == "Women"  & ChildStatus == "No children" ~ "#F48FB1", # pink
      Gender == "Women"  & ChildStatus == "Has children" ~ "#E53935"  # red
    ),
    who = paste(if_else(Gender=="Men","men","women"),
                if_else(ChildStatus=="Has children","\nwith\nchildren","\nwithout\nchildren")),
    # --- NEW: small horizontal nudge so the two women's labels don't overlap ---
    x_nudge = case_when(
      Gender == "Women" & ChildStatus == "Has children"  ~ -0.012, # move a bit left
      Gender == "Women" & ChildStatus == "No children"   ~  0.012, # move a bit right
      TRUE ~ 0
    )
  )

# handy values for brackets
g_men_nc  <- pts$gap[pts$Gender=="Men"   & pts$ChildStatus=="No children"]
g_men_wc  <- pts$gap[pts$Gender=="Men"   & pts$ChildStatus=="Has children"]
g_wom_nc  <- pts$gap[pts$Gender=="Women" & pts$ChildStatus=="No children"]
g_wom_wc  <- pts$gap[pts$Gender=="Women" & pts$ChildStatus=="Has children"]

# differences to show above brackets
delta_womNC_menNC <- g_men_nc - g_wom_nc
delta_womWC_menWC <- g_men_wc - g_wom_wc

# number-line limits
x_left  <- -0.20
x_right <-  0.20

# y positions
y_line  <- 0
y_pct   <- -0.03     # percent just under each dot
y_who   <- -0.10     # two-line labels under the percent
y_brk1  <-  0.12     # small bracket height
y_brk2  <-  0.20     # big bracket height

# a small helper to draw squared brackets
bracket <- function(x0, x1, y, tick = 0.03, ...) {
  list(
    annotate("segment", x = x0, xend = x0, y = y, yend = y - tick, ...),
    annotate("segment", x = x0, xend = x1, y = y, yend = y, ...),
    annotate("segment", x = x1, xend = x1, y = y, yend = y - tick, ...)
  )
}

# Plot
ggplot() +
  # arrowed number line
  geom_segment(aes(x = x_left, xend = x_right, y = y_line, yend = y_line),
               linewidth = 1.2, color = "grey30",
               arrow = arrow(ends = "both", type = "closed", length = unit(6, "pt"))) +
  annotate("text", x = x_left,  y = y_line, label = "Less",  hjust = 1.2, vjust = 0.4, size = 4) +
  annotate("text", x = x_right, y = y_line, label = "More",  hjust = -0.2, vjust = 0.4, size = 4) +
  annotate("text", x = 0, y = y_line - 0.185, label = "Wage comparing to man without children", fontface = "bold", size = 4) +
  # zero tick
  geom_segment(aes(x = 0, xend = 0, y = y_line - 0.04, yend = y_line + 0.04),
               color = "grey45", linewidth = 0.6) +
  # the four dots
  geom_point(data = pts, aes(x = gap, y = y_line), color = pts$dot_col, size = 4) +
  # % under each dot (colored like the dot) — now use x = gap + x_nudge
  geom_text(data = pts,
            aes(x = gap + x_nudge, y = y_pct, label = percent(gap, accuracy = 0.1),
                colour = I(dot_col)),
            size = 3, vjust = 1) +
  # who-labels under the % — now use x = gap + x_nudge
  geom_text(data = pts,
            aes(x = gap + x_nudge, y = y_who, label = who, colour = I(dot_col)),
            lineheight = 0.95, size = 3) +

  # ---- Brackets like your sketch ----
  # small bracket: women w/o children  → men w/o children (≈ 13.9%)
  bracket(g_wom_nc, g_men_nc, y = y_brk1, color = "grey40", linewidth = 0.7)+
  annotate("text", x = (g_wom_nc + g_men_nc)/2, y = y_brk1 + 0.03,
           label = percent(delta_womNC_menNC, accuracy = 0.1), size = 4)+
  # big bracket: women w/ children → men w/ children (≈ 22.2%)
  bracket(g_wom_wc, g_men_wc, y = y_brk2, color = "black", linewidth = 0.7)+
  annotate("text", x = (g_wom_wc + g_men_wc)/2, y = y_brk2 + 0.03,
           label = percent(delta_womWC_menWC, accuracy = 0.1), size = 4)+

  # axis & theme
  scale_x_continuous(labels = percent_format(accuracy = 10),
                     breaks = seq(-0.2, 0.2, by = 0.1)) +
  coord_cartesian(xlim = c(x_left, x_right), ylim = c(-0.26, 0.28), clip = "off") +
  labs(
    title = "How parenthood shapes the gender wage gap relative to men without children\n(ages 25–44)",
    caption = "Top line: wage gaps for men (with vs. without children); bottom line: wage gaps for women relative to men without \nchildren. Parenthood is associated with a wage premium for men and a pronounced penalty for women—largest \nfor women with children.",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(t = 6, r = 12, b = 8, l = 12),
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 10.2),
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0),
  )


```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
another <- read_parquet("./data/01-cleaned_data/cleaned_unemployment_data.parquet")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
library(tidyverse)
library(scales)
library(tidyverse)
library(scales)

## 1) Prep from `another`
lf <- another %>%
  # keep only the labour force rows (your file is already filtered, but this is harmless)
  filter(Employment_status %in% c("Employed", "Unemployed")) %>%
  select(Age, Gender, Unemployed)

## 2) Unemployment rate by age × gender
age_gender_rate <- lf %>%
  group_by(Age, Gender) %>%
  summarise(
    unemployed = sum(Unemployed, na.rm = TRUE),
    labour     = n(),
    rate       = unemployed / labour,
    .groups    = "drop"
  )

## 3) Plot — Men blue, Women red
ggplot(age_gender_rate,
       aes(x = Age, y = rate, group = Gender, color = Gender)) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c("Men" = "#4F83FF", "Women" = "#E15759")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1), limits = c(0, NA)) +
  labs(
    title = "Unemployment rate by age",
    subtitle = "Among the labour force (employed + unemployed), ages 25–64",
    x = "Age group",
    y = "Unemployment rate",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = "top",
    panel.grid.minor  = element_blank()
  )

```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
library(tidyverse)
library(scales)

# Expecting clean_unemp with columns:
# Age (factor), Gender ("Women"/"Men"), Employment_status, Unemployed (logical), ChildStatus ("Has children"/"No children")
# If you loaded it as `another`, just do: clean_unemp <- another

# Keep women in the labour force (your file already filtered to InLF, but this is harmless)
lf_women <- another %>%
  filter(Gender == "Women",
         Employment_status %in% c("Employed","Unemployed"),
         !is.na(ChildStatus))

# Unemployment rate by Age × ChildStatus
urate_women_age <- lf_women %>%
  group_by(Age, ChildStatus) %>%
  summarise(
    labour = n(),
    urate  = mean(Unemployed, na.rm = TRUE),   # Unemployed / (Employed + Unemployed)
    .groups = "drop"
  )

# Plot: Has children = red, No children = pink
ggplot(urate_women_age, aes(x = Age, y = urate, color = ChildStatus, group = ChildStatus)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Has children" = "#E53935",  # red
                                "No children" = "#F48FB1")) + # pink  ← plus goes here
  scale_y_continuous(labels = percent_format(accuracy = 0.1), limits = c(0, NA)) +
  labs(
    title = "Unemployment rate by age among women: with vs. without children",
    subtitle = "Rate = Unemployed / (Employed + Unemployed) within each age group",
    x = "Age group", y = "Unemployment rate", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```


\newpage


# References
