---
title: "The Motherhood Penalty: Why Gender Equality at Work Still Eludes Canada"
subtitle: "An analysis of how childbirth reshapes women’s earnings, employment, and opportunity"
author:
  - "Sakura Hu"
  - "Zhanyi Wang"
date: "`r format(as.Date('2025-10-24'), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(arrow)
library(readr)
library(scales)
library(ggrepel)
library(forcats)
library(dplyr)
library(tidyr)
library(ggplot2)
library(grid)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data <- read_parquet("./data/01-cleaned_data/cleaned_census_data.parquet")
```
# Persistent Gender Inequality in the Labour Market 
In recent years, in order to address problems such as gender discrimination and the undervaluation of women’s work, many countries and regions promote equal pay for work of equal value and the economic empowerment of females[@one]. As one of these regions, Canada has made continuous progress in addressing gender inequality in employment, but the pay gap remains stubbornly persistent. Women with full-time jobs still earn significantly less than men on average, and this gap is even wider for Indigenous women, women of colour, women with disabilities, and others who face systemic and intersectional barriers[@two]. In this context, an important question arises: Why does the gender pay gap persist even after policies like equal pay for equal work have been implemented? Education and occupation are often cited as explanations, but the data tells a more complex story, which relates the wage, gender, and motherhood.

# Education and Occupation Alone Are Not Enough To Explain the Wage Gap 
A common hypothesis attributes wage differences to education and occupation gaps. In general, higher education means a greater chance of entering occupations with higher average salaries[@three], so if most women choose to work early instead of continuing to obtain higher education, they may lose the opportunity to access high-paying occupations. However, this explanation no longer holds. In fact, women are more likely than men to pursue postsecondary studies, and they represent the majority of bachelor’s degree, college certificate and diploma holders[@four]. In addition, across all education levels and occupations, women’s earnings are always lower than men’s. Figure 1 shows that the gender pay gap varies widely by industry, ranging from 9.1% in Utilities to 30% in Finance and Insurance. Sectors such as Professional, Scientific and Technical Services (28%) and Agriculture, Forestry, Fishing and Hunting (26.2%) also exhibit large disparities, indicating that gender inequality persists even in high-skilled and specialized occupations. Figure 2 illustrates the gender pay gap by education level. While the gap narrows with higher education, it does not disappear. Women with a bachelor’s degree or higher still earn about 16.8% less than men, and the largest gap—25.3%—occurs among those with non-degree postsecondary credentials. These results suggest that although higher education and occupational advancement can reduce the gender wage gap, they are insufficient to eliminate it entirely.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# gap by occupation

# 2-digit NAICS sector labels
naics_labs <- c(
  `11` = "Agriculture, forestry, fishing & hunting",
  `21` = "Mining, quarrying, oil & gas",
  `22` = "Utilities",
  `23` = "Construction",
  `31` = "Manufacturing",
  `41` = "Wholesale trade",
  `44` = "Retail trade",
  `48` = "Transportation & warehousing",
  `51` = "Information & culture",
  `52` = "Finance & insurance",
  `53` = "Real estate, rental & leasing",
  `54` = "Professional, scientific & technical",
  `55` = "Management of companies",
  `56` = "Admin & support, waste mgmt",
  `61` = "Educational services",
  `62` = "Health care & social assistance",
  `71` = "Arts, entertainment & recreation",
  `72` = "Accommodation & food services",
  `81` = "Other services (except public admin)",
  `91` = "Public administration"
)


naics_gap <- data %>%
  mutate(NAICS_lab = recode(as.character(NAICS), !!!naics_labs)) %>%
  filter(!is.na(NAICS_lab)) %>%
  group_by(NAICS_lab, Gender) %>%
  summarise(med_wage = median(Wages, na.rm = TRUE), .groups = "drop_last") %>%
  pivot_wider(names_from = Gender, values_from = med_wage) %>%
  mutate(
    gap_pct = (Men - Women) / Men,
    # (Men − Women) / Men
    gap_cat = case_when(
      gap_pct < 0    ~ "Women higher",
      gap_pct < 0.10 ~ "0–10%",
      gap_pct < 0.20 ~ "10–20%",
      TRUE           ~ "≥20%"
    ),
    # set factor order HIGH → LOW so bars run top (largest) to bottom (smallest) after coord_flip()
    NAICS_lab = fct_reorder(NAICS_lab, gap_pct, .desc = TRUE)
  ) %>%
  tidyr::drop_na(gap_pct)

# palette with strong separation (green for negative, oranges/reds for larger gaps)
pal <- c(
  "Women higher" = "#2ca25f",
  "0–10%"       = "#fee8c8",
  "10–20%"      = "#fdbb84",
  "≥20%"        = "#e34a33"
)

# --- single plot object ---
p_naics <- ggplot(naics_gap, aes(
  x = reorder(NAICS_lab, gap_pct),
  y = gap_pct,
  fill = gap_cat
)) +
  geom_col(width = 0.72) +
  geom_text(
    aes(label = percent(gap_pct, accuracy = 0.1)),
    hjust = -0.12,
    size = 3.1,
    colour = "grey20"
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(name = "Gap size", values = pal) +
  labs(
    title   = "How does gender pay gap differ by industy?",
    x       = NULL,
    y       = "Gap percentage",
    caption = "Figure 1: the gender pay gap by industry for full-time, full-year workers. Gaps exist across all sectors, ranging \nroughly from 9% to 30%. Finance & insurance is among the highest."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor    = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption.position = "plot",
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

p_naics
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# gap by education
# Compute median wages by Education × Gender, then the gap = (Men − Women) / Men
edu_gap <- data %>%
  filter(!is.na(Edu3)) %>%
  group_by(Edu3, Gender) %>%
  summarise(med_wage = median(Wages, na.rm = TRUE), .groups = "drop_last") %>%
  pivot_wider(names_from = Gender, values_from = med_wage) %>%
  mutate(gap_pct = (Men - Women) / Men) %>%
  drop_na(gap_pct) %>%
  # order categories from lowest → highest gap
  mutate(Edu3 = as.character(Edu3)) %>%
  arrange(gap_pct) %>%
  mutate(Edu3 = factor(Edu3, levels = Edu3))

# palette per education level (tweak if you like)
edu_cols <- c(
  "HS or less"          = "#fdbb84",
  "Non-degree postsec"  = "#9ecae1",
  "BA or higher"        = "#a1d99b"
)

ggplot(edu_gap, aes(x = gap_pct, y = Edu3, fill = Edu3)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             colour = "grey55") +
  geom_text(
    aes(label = percent(gap_pct, accuracy = 0.1)),
    hjust = -0.12,
    size = 3.2,
    colour = "grey20"
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = edu_cols, name = "Education") +
  labs(
    title   = "How does education level effect gender pay gap?",
    x       = "Gap percentage",
    y       = NULL,
    caption = "Figure 2: the gender pay gap by education level. Although gap exist in all education level, it is largest among those \nwith non-degree postsecondary credentials (25.3%) and smallest among those with a bachelor’s degree or \nhigher (16.8%)."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "right",
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption.position = "plot",
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

```
If education and occupation cannot fully explain the difference, what does?  The answer might be found in a more personal and common moment ---- childbirth.

# Turning point at age 30
When plotting the wage gap between men and women across age, a distinct pattern appears. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
#2021 gap by age

# Map census AGEGRP codes to labels
age_labels <- c(
  `7` = "18–19",
  `8` = "20–24",
  `9` = "25–29",
  `10` = "30–34",
  `11` = "35–39",
  `12` = "40–44",
  `13` = "45–49",
  `14` = "50–54",
  `15` = "55–59",
  `16` = "60–64"
)

# Prep 2021 data
df21 <- data %>%
  mutate(
    Gender = recode(as.character(Gender), `1` = "Women", `2` = "Men"),
    Age = recode(as.character(AGEGRP), !!!age_labels)
  ) %>%
  filter(Gender %in% c("Women", "Men"), !is.na(Wages), !is.na(Age))

# Compute gap by age for 2021: Gap = (Men − Women) / Men using medians
gap_age <- df21 %>%
  group_by(Age, Gender) %>%
  summarise(med = median(Wages, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Gender, values_from = med) %>%
  filter(!is.na(Men), !is.na(Women)) %>%
  mutate(Age = factor(Age, levels = age_labels),
         gap_pct = (Men - Women) / Men) %>%
  arrange(Age)

# Red highlight point at 25–29
pts_2529     <- gap_age %>% filter(Age == "25–29")
pt_2529_top  <- pts_2529  # used for the ggrepel note

p_2021 <- ggplot(gap_age, aes(Age, gap_pct, group = 1)) +
  geom_line(linewidth = 1.1, color = "black") +
  geom_point(size = 2, color = "black") +
  # red dot + label at 25–29
  geom_point(
    data = pts_2529,
    aes(Age, gap_pct),
    inherit.aes = FALSE,
    size = 3.8,
    color = "red"
  ) +
  geom_text_repel(
    data = pts_2529,
    aes(Age, gap_pct, label = "Mean maternal \nage in Canada)"),
    color = "red",
    segment.color = "red",
    nudge_y = 0.06,
    box.padding = 0.35,
    size = 3.2,
    show.legend = FALSE
  ) +
  # Y axis: 0–30%
  scale_y_continuous(
    limits = c(0, 0.30),
    breaks = seq(0, 0.30, by = 0.05),
    labels = scales::label_percent(accuracy = 1)
  ) +
  labs(
    title = "The Turning Point: How the Gender Pay Gap \nExpands Around the Childbearing Years",
    x = "Age group",
    y = "Gap percentage",
    caption = "Figure 3: the gender pay gap by age for full-time, full-year workers in Canada(2021). The gap narrows in the \nearly 20s, then rises sharply from 25–30 and continues to increase steadily thereafter."
    
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    # center the title
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

p_2021

```

As shown in Figure 3, the gender pay gap narrows during the early stages of workers’ careers, reaching its smallest point around ages 25–29. However, immediately after age 30, the gap begins to widen sharply and continues to increase steadily thereafter. Notably, this turning point aligns closely with the average maternal age in Canada. This suggests that the transition into parenthood may be a key factor influencing women’s income trajectories. Rather than reflecting a lack of skill or ambition, the widening gap likely results from structural and social factors—such as career interruptions, reduced working hours, or employer biases—that disproportionately affect women once they become mothers. After becoming mothers, many women also tend to shift part of their focus toward childcare and family responsibilities, which may further reduce their opportunities for promotion or salary growth.

It is also important to recognize that childbirth affects women not only through changes in wages, but also through changes in labour participation itself. After becoming mothers, some women temporarily or permanently leave the workforce to take on domestic and caregiving responsibilities. These women are often classified as “not employed”. Therefore, to fully understand the impact of motherhood on economic inequality, it is necessary to examine both groups, women who remain in paid employment and women who exit the labour market after childbirth. By considering both wage differences among working women and unemployment patterns related to family responsibilities, it can capture a more complete picture of how motherhood affects the outcomes of women's career and economies.

# Parenthood’s Unequal Impact on Men’s and Women’s Careers 
Among employed individuals, the gender gap in earnings widens sharply after parenthood. As shown in Figure 4., among workers without children, women earn about 13% less than men, but after having children, the gap expands to roughly 21%. This pattern suggests that while a gender pay gap already exists before parenthood, it becomes substantially larger once children enter the picture, which suggests that having children can exacerbate pay inequality in the workplace.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
# --- Parent premium/penalty on a single number line --------------------------

pts <- data %>%
  mutate(
    Gender = recode(as.character(Gender), `1`="Women", `2`="Men"),
    ChildStatus = case_when(PKIDS == 1 ~ "Has children",
                            PKIDS == 0 ~ "No children",
                            TRUE ~ NA_character_)
  ) %>%
  filter(AGEGRP %in% 9:12,                      # 25–44
         Gender %in% c("Men","Women"),
         !is.na(ChildStatus),
         !is.na(Wages)) %>%
  group_by(Gender, ChildStatus) %>%
  summarise(med = median(Wages, na.rm = TRUE), .groups = "drop") %>%
  # baseline = Men — No children
  mutate(baseline = med[Gender == "Men" & ChildStatus == "No children"][1],
         gap = (med - baseline) / baseline) %>%
  mutate(
    # aesthetics for the four dots
    dot_col = case_when(
      Gender == "Men"    & ChildStatus == "No children" ~ "#4F83FF", # blue (baseline)
      Gender == "Men"    & ChildStatus == "Has children" ~ "#2E5BD4", # darker blue
      Gender == "Women"  & ChildStatus == "No children" ~ "#F48FB1", # pink
      Gender == "Women"  & ChildStatus == "Has children" ~ "#E53935"  # red
    ),
    who = paste(if_else(Gender=="Men","men","women"),
                if_else(ChildStatus=="Has children","\nwith\nchildren","\nwithout\nchildren")),
    # --- NEW: small horizontal nudge so the two women's labels don't overlap ---
    x_nudge = case_when(
      Gender == "Women" & ChildStatus == "Has children"  ~ -0.012, # move a bit left
      Gender == "Women" & ChildStatus == "No children"   ~  0.012, # move a bit right
      TRUE ~ 0
    )
  )

# handy values for brackets
g_men_nc  <- pts$gap[pts$Gender=="Men"   & pts$ChildStatus=="No children"]
g_men_wc  <- pts$gap[pts$Gender=="Men"   & pts$ChildStatus=="Has children"]
g_wom_nc  <- pts$gap[pts$Gender=="Women" & pts$ChildStatus=="No children"]
g_wom_wc  <- pts$gap[pts$Gender=="Women" & pts$ChildStatus=="Has children"]

# differences to show above brackets
delta_womNC_menNC <- g_men_nc - g_wom_nc
delta_womWC_menWC <- g_men_wc - g_wom_wc

# number-line limits
x_left  <- -0.20
x_right <-  0.20

# y positions
y_line  <- 0
y_pct   <- -0.03     # percent just under each dot
y_who   <- -0.10     # two-line labels under the percent
y_brk1  <-  0.12     # small bracket height
y_brk2  <-  0.20     # big bracket height

# a small helper to draw squared brackets
bracket <- function(x0, x1, y, tick = 0.03, ...) {
  list(
    annotate("segment", x = x0, xend = x0, y = y, yend = y - tick, ...),
    annotate("segment", x = x0, xend = x1, y = y, yend = y, ...),
    annotate("segment", x = x1, xend = x1, y = y, yend = y - tick, ...)
  )
}

# Plot
ggplot() +
  # arrowed number line
  geom_segment(aes(x = x_left, xend = x_right, y = y_line, yend = y_line),
               linewidth = 1.2, color = "grey30",
               arrow = arrow(ends = "both", type = "closed", length = unit(6, "pt"))) +
  annotate("text", x = x_left,  y = y_line, label = "Less",  hjust = 1.2, vjust = 0.4, size = 4) +
  annotate("text", x = x_right, y = y_line, label = "More",  hjust = -0.2, vjust = 0.4, size = 4) +
  annotate("text", x = 0, y = y_line - 0.185, label = "Wage comparing to man without children", fontface = "bold", size = 4) +
  # zero tick
  geom_segment(aes(x = 0, xend = 0, y = y_line - 0.04, yend = y_line + 0.04),
               color = "grey45", linewidth = 0.6) +
  # the four dots
  geom_point(data = pts, aes(x = gap, y = y_line), color = pts$dot_col, size = 4) +
  # % under each dot (colored like the dot) — now use x = gap + x_nudge
  geom_text(data = pts,
            aes(x = gap + x_nudge, y = y_pct, label = percent(gap, accuracy = 0.1),
                colour = I(dot_col)),
            size = 3, vjust = 1) +
  # who-labels under the % — now use x = gap + x_nudge
  geom_text(data = pts,
            aes(x = gap + x_nudge, y = y_who, label = who, colour = I(dot_col)),
            lineheight = 0.95, size = 3) +

  # ---- Brackets like your sketch ----
  # small bracket: women w/o children  → men w/o children (≈ 13.9%)
  bracket(g_wom_nc, g_men_nc, y = y_brk1, color = "grey40", linewidth = 0.7)+
  annotate("text", x = (g_wom_nc + g_men_nc)/2, y = y_brk1 + 0.03,
           label = percent(delta_womNC_menNC, accuracy = 0.1), size = 4)+
  # big bracket: women w/ children → men w/ children (≈ 22.2%)
  bracket(g_wom_wc, g_men_wc, y = y_brk2, color = "black", linewidth = 0.7)+
  annotate("text", x = (g_wom_wc + g_men_wc)/2, y = y_brk2 + 0.03,
           label = percent(delta_womWC_menWC, accuracy = 0.1), size = 4)+

  # axis & theme
  scale_x_continuous(labels = percent_format(accuracy = 10),
                     breaks = seq(-0.2, 0.2, by = 0.1)) +
  coord_cartesian(xlim = c(x_left, x_right), ylim = c(-0.26, 0.28), clip = "off") +
  labs(
    title = "How Having Children Alters Earnings for Men and Women?",
    caption = "Figure 4: Wage differences by parenthood status for men and women aged 25–44, relative to men \nwithout children. Men with children earn about 6.9% more than men without children, while women \nwith children earn about 15.3% less. This results in an overall gender wage gap of 22.2% among \nparents, and 13.9% among non-parents.",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(t = 6, r = 12, b = 8, l = 12),
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 10.2),
    plot.caption.position = "plot",
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

```

Furthermore, the impact of having children on earnings moves in opposite directions for men and women. Men with children earn about 7% more than men without children, as many employers tend to view fathers as more stable, responsible, and committed. These qualities can translate into higher pay and faster promotions[@five].

In contrast with men, women without children earn about 14% less than men without children, but after having children, women earn around 15% less. The increasing wage gap after having a child may reflect cultural biases toward mothers, which to be seen as a good mother they are expected to prioritize family over career, but to be viewed as a “Good leader,” they must do the opposite. Under such contradiction, many mothers deviate from traditional career paths by reducing their working hours or leaving the workforce indefinitely, all of which negatively affects their chances of career advancement[@eight]. The starkly different career situations of women and men after having children illustrate that gender stereotypes and unfairness in the workplace still exist, exacerbating the gender wage gap.

# Childbirth and Women’s Employment Patterns
While the gender pay gap widens around the mean maternal age, the impact of childbirth is also reflected in women’s unemployment patterns. As shown in Figure 5, women with children consistently face higher unemployment rates than women without children across all age groups, particularly between ages 25 and 45. This is because, in addition to childbirth, women in this age group are more likely to take on caregiving responsibilities for children under the age of 15 [@six]. In contrast, women without children experience a much sharper decline in unemployment after their early twenties and maintain relatively stable employment throughout midlife.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
another <- read_parquet("./data/01-cleaned_data/cleaned_unemployment_data.parquet")
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}

lf_women <- another %>%
  filter(Gender == "Women",
         Employment_status %in% c("Employed","Unemployed"),
         !is.na(ChildStatus))

# Unemployment rate by Age × ChildStatus
urate_women_age <- lf_women %>%
  group_by(Age, ChildStatus) %>%
  summarise(
    labour = n(),
    urate  = mean(Unemployed, na.rm = TRUE),   # Unemployed / (Employed + Unemployed)
    .groups = "drop"
  )

# Plot: Has children = red, No children = pink
ggplot(urate_women_age, aes(x = Age, y = urate, color = ChildStatus, group = ChildStatus)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Has children" = "#E53935",  # red
                                "No children" = "#F48FB1")) + # pink  ← plus goes here
  scale_y_continuous(labels = percent_format(accuracy = 0.1), limits = c(0, NA)) +
  labs(
    title = "How Parenthood Shapes Women’s Employment \nAcross Age Groups?",
    caption = "Figure 5: Compares unemployment rates among women with and without children across age groups. Women with \nchildren face higher unemployment rates, particularly between ages 25 and 45 — the main childbearing and \ncaregiving years.",
    x = "Age group", y = "Unemployment rate", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank(),
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 10.2),
    plot.caption.position = "plot",
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )
```

This pattern suggests that childbirth can interrupt women’s participation in the labour market. Many women temporarily leave their jobs or reduce their working hours to care for children, and re-entering the workforce often means facing less favorable job conditions.

When viewed alongside Figure 6, which compares unemployment patterns between men and women, the impact of gender and parenthood becomes even clearer. Although men and women show similar unemployment rates at younger and older ages, the gap between male and female unemployment widens noticeably between ages 29 and 40, which corresponds to the main childbirth and childrearing periods. This pattern indicates that age and gender do not operate independently, and the effect of gender on unemployment becomes much stronger during the childrearing years. In other words, women experience a temporary rise in unemployment precisely when family responsibilities are most demanding, while men’s employment remains relatively stable across all ages.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
## 1) Prep from `another`
lf <- another %>%
  # keep only the labour force rows (your file is already filtered, but this is harmless)
  filter(Employment_status %in% c("Employed", "Unemployed")) %>%
  select(Age, Gender, Unemployed)

## 2) Unemployment rate by age × gender
age_gender_rate <- lf %>%
  group_by(Age, Gender) %>%
  summarise(
    unemployed = sum(Unemployed, na.rm = TRUE),
    labour     = n(),
    rate       = unemployed / labour,
    .groups    = "drop"
  )

## 3) Plot — Men blue, Women red
ggplot(age_gender_rate,
       aes(x = Age, y = rate, group = Gender, color = Gender)) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c("Men" = "#4F83FF", "Women" = "#E15759")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1), limits = c(0, NA)) +
  labs(
    title = "Gender Differences in Employment Across Age Groups",
    caption = "Figure 6: compares the unemployment rate for men and women across different age groups. Men and women have \nsimilar unemployment patterns at younger and older ages, but between ages 30 and 40, women face higher \nunemployment rates, corresponding to the main childbearing years.",
    x = "Age group",
    y = "Unemployment rate",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = "top",
    panel.grid.minor  = element_blank(),
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 10.2),
    plot.caption.position = "plot",
    plot.caption = element_text(
      size = 9.5,
      colour = "#5a5a5a",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

```

# Limitations and Future Research Directions
Despite extensive analysis and research, the article still has certain limitations. For example, when considering women’s employment status, only full-time and unemployed women were included, but many women engage in part-time work after childbirth, which was not considered in the study.Moreover, since the data come from census results, outcomes such as unemployment rates are influenced by more than just childbirth. Other social and economic factors that were not included in the study may also affect the results. For instance, an economic downturn or slow growth can lead to higher unemployment rates, which may disproportionately impact certain industries or regions[@seven]. In such cases, the observed differences between men’s and women’s employment might reflect broader economic conditions rather than the direct effect of childbirth. In addition, some women may face anticipatory discrimination during childbearing age even if they do not have children, because employers expect future maternity leave or reduced commitment. These unobserved factors can affect wages and employment independently of actual childbirth and may create gaps between the findings and real conditions.

In order to solve the impact of the above limitations, more data can be associated in future research. First, employment data can be combined with census data to include part-time women workers. This would help show a complete picture of women’s work patterns after having children. Then, the stratified analysis can also be applied to look at the proportion of women in senior and management positions and the pay gap. This approach can help identify whether gender inequality becomes more pronounced at higher career levels. It can also reveal if women face barriers to promotion or slower wage growth compared to men in similar positions. Including these methods and data can make the research results more realistic.

By addressing these research gaps, future studies can provide stronger evidence for policymaking. Based on the current findings, some practical social actions can also be taken to reduce gender inequality in the workplace. For example, flexible and family-friendly work options, such as hybrid work, remote work, and part-time leadership jobs, can help parents balance work and family without losing chances for promotion. More affordable and high-quality childcare services would also help women stay in or return to full-time jobs after having children. The government can support this by giving more funding and subsidies to childcare programs.

# Rethinking Equality Beyond Pay
In today’s workforce, education, ambition, and skill are no longer the main differences between men and women. The change usually happens around age 30, which is the age many people become parents. The career gap between men and women, which narrows in the early stages of their careers, widens again as family responsibilities increase.

Thus, the gender pay gap is not merely an economic measure; it reflects deeper societal attitudes toward the division of labour at home and at work. Closing this gap requires more than equal pay legislation — it demands structural changes that address the employment inequalities women face due to childbirth, caregiving, and related constraints. Only by fostering a truly supportive and inclusive work environment can we move closer to genuine gender equality.

\newpage


# References
